---
title: "Pair Programming R Workshop Tasks"
format: html
---

# Pair Programming Exercises 

In the following exercises, you will be working with your partner(s) through a series of exercises demonstrating how to write functions in R.

Do not worry if you do not get through all of the exercises. Switch the driver and navigator(s) roles after each exercise.

# Task 1 
## 1.1 Warmup (what are functions)

**Solved** Exercise 1.1: Functions in R are 'code shortcuts' to common things you want to do.

Functions 'take' some inputs/variables in round brackets; then perform some operations on those inputs in `{ }` and then return one value with `return(...)`. Explicit returns by calling the `return()` function are optional/stylistic, you could also call the result object. 

```{r}
#| label: pseudo-code-function
#| eval: false 

add <- function( first_argument, second_argument ) {
  result <- first_argument + second_argument
  return(result)
}
```

Once defined, you can use your function ('run it') many times with different inputs. This saves you copy-pasting code.

```{r}
#| label: pseudo-code-function-use
#| eval: false 

result1 <- add( 4, 6 )
result2 <- add( 9, 7 )
result3 <- add( result1, result2 )
```

Read the following code and discuss with your partner(s) what it does.

```{r}
#| label: exercise1-1a

# Function 1: number_of_full_dozens 

number_of_full_dozens <- function(number_of_eggs) {
  full_dozens <- number_of_eggs %/% 12 # %/% means "divide and keep integer"
  return(full_dozens)
}

print("below should return: 0, 1, 2, 8")
number_of_full_dozens(11) # 0
number_of_full_dozens(12) # 1 
number_of_full_dozens(25) # 2 
number_of_full_dozens(100) # 8 
```

```{r}
#| label: exercise1-1b

# Function 2: is_first_larger 

is_first_larger <- function( first, second ) {
  return(first > second)
}

print("below should return: FALSE, TRUE, FALSE")
is_first_larger(11,12) # FALSE
is_first_larger(12,11) # TRUE
is_first_larger(12,12) # FALSE 
```

> Switch who is driver now! 

## 1.2: Unit conversion: Temperature 

Create a function called `convert_temperature_CtoF` that takes a numeric value (a temperature in Celsius) as the argument, and returns the temperature in Fahrenheit. 

To get Fahrenheit you need to multiply Celsius by 9, then divide by 5 and then add 32.

e.g., 10 Celsius is 50 Fahrenheit because 10 * 9 / 5 + 32 = 50

Return values rounded with `round()`

```{r}
#| label: exercise1-2 

convert_temp_CtoF <- function( something ) {
  result <- something + something
  return(result)
}

print("below should return: 32, 50, 99, 212")
convert_temp_CtoF(0) # 32
convert_temp_CtoF(10) # 50 
convert_temp_CtoF(37) # 99 
convert_temp_CtoF(100) # 212 
```

> Switch who is driver now! 

## 1.3 Unit Conversion: Temperature 

Now create a function for conversion the other way called `convert_temp_FtoC` that takes a numeric value (a temperature in Fahrenheit) as the argument.

To get this result, just flip the math from previous task. You can do it!

Return values rounded with `round()`

```{r}
#| label: exercise1-3 

# your code here


print("below should return: -18, -12, 3, 38")
convert_temp_FtoC(0) # -18 
convert_temp_FtoC(10) # -12
convert_temp_FtoC(37) # 3 
convert_temp_FtoC(100) # 38 
```

> Switch who is driver now! 

## 1.4 Unit Conversion: Measurement 

Using your solutions above as a guide, create a function called `convert_length_INtoCM` that takes a numeric value (a unit of length in inches) as the argument.

**HINT:** the inches to cm formula: cm = inches * 2.54

```{r}
#| label: exercise1-4 


# your code here


print("below should return: 3, 10, 25, 99")
convert_length_INtoCM(1) # 3
convert_length_INtoCM(4) # 10 
convert_length_INtoCM(10) # 25
convert_length_INtoCM(39) # 99 
```

> Switch who is driver now! 

# Task 2 
## 2.1 Functions with Multiple Arguments 

Create a function called `which_is_larger` which determines which of the 2 numeric inputs is larger.

Return a string describing what happened. See tests for the exact format of expected outputs.

```{r}
#| label: exercise2-1

which_is_larger <- function ( give_this_better_name , and_this_too ){
  if( ... )
    return(...)
  else if(...)
    return(...)
  else 
    return(...)
}

print("below should return strings: 'second', 'first', 'they're the same!'")
which_is_larger(4, 5) # second 
which_is_larger(5, 4) # first 
which_is_larger(5, 5) # they're the same! 
```

> Switch who is driver now! 

## 2.2  Functions with Multiple Arguments 

Create a function called `which_is_largest` that checks 3 numbers and tells you which one is largest by returning a character string "first", "second" or "third".

If there are ANY two identical numbers, just return a string "oh no!"

```{r}
#| label: exercise2-2

which_is_largest <- function ( first, second, third ){
  return("banana")
}

print("below should return strings: 'oh no!', 'oh no!', 'oh no!'")
which_is_largest(5, 5, 6) # oh no 
which_is_largest(5, 6, 6) # oh no 
which_is_largest(6, 6, 6) # oh no 

print("below should return strings: 'third', 'second', 'first'")
which_is_largest(4, 5, 6) # third
which_is_largest(4, 50, 6) # second
which_is_largest(40, 5, 6) # first

```

> Switch who is driver now! 

# Task 3 
## 3.1: Functions and Vectorisation

Complete the two example functions `twice` and `is_positive` in the code chunks below 

```{r}
#| label: exercise3-1a 

# Function 1: twice 

twice <- function(num){
  return(....) 
}

print("below should return: 4, 88")
twice(2) # 4 
twice(44) # 88 
```

```{r}
#| label: exercise3-1b

# Function 2: is_positive 
is_positive <- function(num){
  if(....){
    return(....)
  } else{
    return(....)
  }
} 
 
print("below should return: FALSE, TRUE")
is_positive(-20) # FALSE
is_positive(50) # TRUE
```

> Switch who is driver now! 

## 3.2: Functions and Vectorisation

Line by line, uncomment the tests **ONE AT THE TIME** and see what happens. Discuss with your partner(s).

```{r}
#| label: exercise3-2a

twice <- function(num){
  return(num *2) 
}

is_positive <- function(num){
  if(num > 0){
    return(TRUE)
  } else{
    return(FALSE)
  }
} 

# uncomment the tests ONE AT THE TIME and see what happens.

values <- c(45, -60, 150, -85, 20)
# twice(values)
# is_positive(values)
```

~ Only after running the tests and exploring the output, read below ~




... oh no, an error! ...

Not to to fear!

Most functions in R are vectorized, meaning that the function will operate on all elements of a vector without needing to loop through and act on each element one at a time.

if else statements are *NOT* vectorized, hence the above error. However, the `*` (multiplication operator) is vectorized, which is why the twice function works with the values vector!

Vectorization is a real superpower of R! It makes writing code more concise, easy to read, and less error prone.

There is a vectorized if else statement in R, `ifelse()`, which we show in the `is_positive_vectorized()` function as well below.

```{r}
#| label: exercise3-2b 

# Vectorized solution 
is_positive_vectorized <- function(num){
  # either
  ifelse(num > 0, TRUE, FALSE)
  # or, even better, because > is vectorized, just
  # num > 0
}

is_positive_vectorized(values)
```

> Switch who is driver now! 

# Task 4 
## 4.1: Functions and Plots 

In this exercise, we will be using a function to programmatically change the colors of a plot. Have a look at the starting code, discuss what you think it does with your partner(s), and move on to the next section.

```{r}
#| label: exercise4-1-starter
 
# Exercise 4.1: starting code.
# Have a read and run it. What does it do?

library(ggplot2)
library(purrr)

fruit_data <- data.frame(
  fruit = c("Apples", "Bananas", "Cherries", "Blueberries", "Oranges"),
  sales = c(100, 150, 8, 58, 90)
)

ggplot(data = fruit_data, aes(x = fruit, 
                              y = sales)) + 
  geom_col()
```

> Switch who is driver now! 

## 4.2: It's your turn now!

Create a function called `categorize_and_color` that takes a number as input and returns a character string with the name of a color (e.g., 'red', 'orange' or 'green') based on some logic (decide on the logic with your partner(s)). It's up to you to come up with the logic rules.

All other code is already written for you - just complete the body of the function `categorize_and_color`.

```{r}
#| label: exercise4-2

# START your code here. Maybe start with some 'testing prints'?

categorize_and_color <- ....

# END your code here.

# and now we will use your function to add a column and then use that in the graph as our color fill aesthetic.
fruit_data <- data.frame(
  fruit = c("Apples", "Bananas", "Cherries", "Blueberries", "Oranges"),
  sales = c(100, 150, 8, 58, 90)
)

fruit_data$sales_color <- map(fruit_data$sales, categorize_and_color)

ggplot(data = fruit_data, aes(x = fruit, 
                              y = sales,
                              fill = sales_color)) + 
  geom_col()
```

----

That's all for now! 
